---
title: "Conduct a hybrid control analysis on a custom dataset using BDB"
author: "Matt Secrest and Isaac Gravestock"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{2. Conduct a hybrid control analysis on a custom dataset using BDB}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knitr, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  fig.align = "center",
  fig.width = 5,
  fig.height = 3,
  dpi = 120,
  comment = "#"
)
options(rmarkdown.html_vignette.check_title = FALSE)
```

In this article, you'll learn how to conduct Bayesian dynamic borrowing (BDB)
analyses on custom data using `psborrow2`.

# Load dependencies and setup CMD Stan

We'll start by loading dependencies. `psborrow2` relies on the package
[`cmdstanr`](https://mc-stan.org/cmdstanr/index.html),
so we'll load [`cmdstanr`](https://mc-stan.org/cmdstanr/index.html)
as well.

```{r dependencies, message = FALSE}
library(psborrow2)
library(cmdstanr)
```

As explained in the
[`cmstanr` vignette](https://mc-stan.org/cmdstanr/articles/cmdstanr.html),
installing the `cmdstanr` R package does not install the underlying CMD Stan
program we need. To install CMD Stan, we will call the `cmdstanr` function
`install_cmdstan()`:

```{r install-cmdstan-2, eval=FALSE}
set_cmdstan_path(path = "/root/.cmdstan")
install_cmdstan(cores = 2)
```

# Creating an analysis object {.tabset}
For an analysis on custom data in `psborrow2`, our goal is to create an object
of class `Analysis` which contains all the information we need to build a model
and compile an MCMC sampler using Stan. To create an `Analysis` object,
we need to call the function `create_analysis_obj()`. Let's look at the four
required arguments to this function and evaluate them one at a time.

```
create_analysis_obj(
  data_matrix,
  outcome,
  borrowing,
  treatment
)
```

## 1. data_matrix {.tabset}

### Required elements

`data_matrix` is where we input the one-row-per-patient data matrix for our
analysis. There are two columns required for all analyses:

* A flag denoting receipt of the experimental intervention (`1`) or not (`0`)
* A flag denoting whether the patient was part of the external data source
(`1`) or the internal trial (`0`).

#### Time to event
If the outcome is time-to-event, then two additional columns are needed:

* The duration of follow-up for each patient
* A flag denoting the patient was censored (`1`) or not (`0`)

#### Binary endpoints
If the outcome is binary, one additional column is needed

* A flag denoting a patient had the event of interest (`1`) or not (`0`)

#### Covariates

Covariates may also be included in BDB analyses. These should be included
in the data matrix if the plan is to adjust for them.

***Note*** R data frames are not allowed, only matrices.

***Note*** No missing data is currently allowed, all values must be non-missing.

### Example data

We will be using an example dataset stored in `psborrow2`
(`psborrow2::example_matrix`). If you are starting from a data frame or tibble,
you can cast to a matrix easily with the `psborrow2` helper function
`create_data_matrix()`, though we don't show that here.

Let's look at the first few rows of the example matrix:

```{r example-matrix}
head(example_matrix)
```

The column definitions are below:

* `ext`, 0/1, flag for external controls
* `trt`, 0/1, flag for treatment arm
* `cov1`, 0/1, baseline covariate
* `cov2`, 0/1, baseline covariate
* `time`, 0/1, numeric >0, survival time
* `cnsr`, 0/1, censoring indicator
* `resp`, 0/1, indicator for binary response outcome

## 2. outcome

`psborrow2` currently supports three outcomes, each of which is created via its
own constructor:

* Time to event with exponential distribution (constant hazard), created with
`exp_surv_dist()`
* Time to event with Weibull distribution and proportional hazards
parametrization, created with `weib_ph_surv_dist()`
* Binary endpoints with a bernoulli distribution (logistic regression),
created with `logistic_bin_outcome()`

For our example, let's conduct a time-to-event analysis using the exponential
distribution.

```{r exp-surv-dist}
outcome <- exp_surv_dist(
  time_var = "time",
  cens_var = "cnsr"
)
outcome
```

## 3. borrowing

`psborrow2` support three different borrowing methods which are specified in
the `method` argument of `borrowing_details()`.

* "No borrowing", this is the internal trial comparison without any external
data
* "Full borrowing", this is pooling of the external and internal control arms
* "BDB", Bayesian dynamic borrowing

In addition to specifying the method, we need to specify a prior distribution
for the baseline event rate, `baseline_prior`. In this case, `baseline_prior` is
a log hazard rate (for binary endpoints, this field represents the log
odds). Let's assume we have no prior knowledge on this event rate, so we'll
specify an uninformative prior: `normal_prior(0, 1000)`.

The column name for the external control column flag in our matrix
is also required and passed to `ext_flag_col`.

Finally, for BDB only,
the hyperprior distribution on the commensurability parameter must be specified.
This hyperprior determines (along with the comparability of the outcomes
between internal and external controls) how much borrowing of the external
control group will be performed. Example hyperpriors include largely
uninformative inverse gamma distributions
e.g., `gamma_prior(alpha = .001, beta = .001)` as well as more
informative distributions e.g., `gamma_prior(alpha = 1, beta = .001`),
though any distribution $x \in (0, \infty)$ can be used. Distributions
with more density at higher values of $x$ (i.e., higher precision)
will lead to more borrowing. We'll choose an uninformative gamma prior
in this example.

***Note***: Prior distributions are outlined in greater detail in a separate
vignette, `vignette('prior_distributions', package = 'psborrow2')`

```{r borrowing-details}
borrowing <- borrowing_details(
  method = "BDB",
  baseline_prior = normal_prior(0, 1000),
  ext_flag_col = "ext",
  tau_prior = gamma_prior(0.001, 0.001)
)
borrowing
```

## 3. treatment

Finally, treatment details are outlined in `treatment_details()`. Here, we first
specify the column for the treatment flag in `trt_flag_col`. In addition, we
need to specify the prior on the effect estimate, `trt_prior`. We'll use
another uninformative normal distribution for the prior on the treatment effect:

```{r treatment-details}
treatment <- treatment_details(
  trt_flag_col = "trt",
  trt_prior = normal_prior(0, 1000)
)
treatment
```

## Application

Now that we have thought through each of the inputs to `create_analysis_obj()`,
let's create an analysis object:

```{r anls-obj}
anls_obj <- create_analysis_obj(
  data_matrix = example_matrix,
  outcome = outcome,
  borrowing = borrowing,
  treatment = treatment
)
```

The Stan model compiled successfully and informed us that we are
ready to begin sampling.

# Sampling from an analysis object

We can take draws from the posterior
distribution using the function `mcmc_sample()`. This function takes as input
our `Analysis` object and any arguments (other than the `data` argument)
that are passed to
[`CmdStanModel` objects](https://mc-stan.org/cmdstanr/reference/model-method-sample.html).

```{r res}
res <- mcmc_sample(anls_obj,
  iter_warmup = 2000,
  iter_sampling = 50000,
  chains = 4,
  seed = 112233
)
```

# Summarizing results{.tabset}

As a `CmdStanMCMC` object, `res` has several methods which are outlined on
the
[`cmdstanr` website](https://mc-stan.org/cmdstanr/reference/CmdStanModel.html).
For instance, we can see a see a summary of the posterior distribution samples
with `res$summary()`:

```{r res-summary}
res$summary()
```

The summary includes information for several parameter estimates
from our BDB model. Because it may not be immediately clear what the parameters
from the Stan model refer to, `psborrow2` has a function which returns a
variable dictionary from the analysis object to help interpret these parameters:

```{r data-dict}
variable_dictionary(anls_obj)
```

We can also capture all of the draws with the function `res$draws()`, which
returns an object of class `draws`. `draws` objects are
common in many MCMC sampling softwares and allow us to leverage
packages such as `posterior` and `bayesplot`.

```{r draws}
draws <- res$draws()
print(draws)
```

***Note*** `psborrow2` also has a function to rename variables in `draws`
objects to be more interpretable, which could be called with e.g.,
`draws <- rename_draws_covariates(draws)`.

## bayesplot

With `draws` objects and `bayesplot`, we can create many useful visual summary
plots. We can visualize the marginal posterior distribution of a
variable we are interested in by plotting histograms of the draws with the
function `mcmc_hist()`. Let's do that for the Hazard ratio for the treatment
effect and for our commensurability parameter, tau.

```{r mcmc-hist, message = FALSE}
library(bayesplot)
mcmc_hist(draws, c("HR_trt", "tau"))
```

We can see other plots that help us understand and diagnose problems with
the MCMC sampler, such as trace and rank plots:

```{r mcmc-trace, fig.width=6, fig.height=3.6, dpi = 100}
color_scheme_set("mix-blue-pink")
mcmc_trace(draws,
  pars = c("HR_trt", "tau"),
  n_warmup = 1000
)
```

Many other functions are outlined in the
[`bayesplot` vignettes](https://mc-stan.org/bayesplot/)

## posterior

`draws` objects are also supported by the `posterior` package,
which provides many other tools for analyzing MCMC draw data. For instance, we
can use `posterior` to derive 80% credible intervals for all parameters:

```{r posterior-summ, message = FALSE}
library(posterior)
summarize_draws(draws, ~ quantile(.x, probs = c(0.1, 0.9)))
```

Another useful application of `posterior` is the evaluation of Monte Carlo
standard error for quantiles of a variable of interest:

```{r mcse-mean}
vm <- extract_variable_matrix(draws, "HR_trt")
mcse_quantile(x = vm, probs = c(0.1, 0.5, 0.9))
```

`posterior` contains many other helpful functions, as outlined on their
[vignettes](https://cran.r-project.org/web/packages/posterior/vignettes/posterior.html).
