---
title: "8. Data Simulation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data_Simulation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options("max.print" = 20)
```

This vignette will show the internal functions for generating data that can be used in simulation studies.

```{r setup}
library(psborrow2)
```

## Generating Baseline Data

First we define how to generate baseline data for our study. In its simplest form we only need to define how many
patients are in the arms: treated internal, control internal, control external.

```{r}
simple_baseline <- create_baseline_object(
  n_trt_int = 100,
  n_ctrl_int = 50,
  n_ctrl_ext = 100
)
```

This object defines how the data will be created. To actually generate a sample, use `generate()` to produce a
`data.frame`.

```{r}
generate(simple_baseline)
```

Often we are interested in some additional covariates. Internally the package uses multivariate normal distributions,
so we need to specify the mean and (co-)variance. We have the option to specify different distribution parameters
for the internal and external arms. The internal arms are assumed to be randomized and therefore come from the same
distribution. If no external parameters are specified, the internal ones are used for both.

```{r}
with_age <- create_baseline_object(
  n_trt_int = 100,
  n_ctrl_int = 50,
  n_ctrl_ext = 100,
  covariates = baseline_covariates(
    names = "age",
    means_int = 55,
    covariance_int = covariance_matrix(5)
  )
)
generate(with_age)
```

In a more complex setting, we can generate correlated baseline covariates.
```{r}
with_age_score <- create_baseline_object(
  n_trt_int = 100,
  n_ctrl_int = 50,
  n_ctrl_ext = 100,
  covariates = baseline_covariates(
    names = c("age", "score"),
    means_int = c(55, 5),
    means_ext = c(60, 5),
    covariance_int = covariance_matrix(c(5, 1))
  )
)
data_age_score <- generate(with_age_score)
```

```{r}
plot(
  x = data_age_score$age,
  y = data_age_score$score,
  col = data_age_score$trt + 1,
  pch = data_age_score$ext * 16,
  xlab = "Age",
  ylab = "Score"
)
legend(
  "topright",
  legend = c("internal treated", "internal control", "external control"),
  col = c(2, 1, 1),
  pch = c(0, 0, 16)
)
```

Finally, we have the option to transform non-normal covariates, such as with binary cut-offs.
These can be added to existing objects and can use built-in functions or used defined.
```{r}
with_age_score@transformations <- list(
  score_high = binary_cutoff("score", int_cutoff = 0.7, ext_cutoff = 0.7),
  score_rounded = function(data) round(data$score)
)
generate(with_age_score)
```
