---
title: "2. Conduct a hybrid control analysis on a dataset using BDB"
author: "Matt Secrest and Isaac Gravestock"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{2. Conduct a hybrid control analysis on a dataset using BDB}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

```{r knitr, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  fig.align = "center",
  fig.width = 5,
  fig.height = 3,
  dpi = 120,
  comment = "#"
)
options(rmarkdown.html_vignette.check_title = FALSE)
```

In this vignette, you'll learn how to conduct Bayesian dynamic borrowing (BDB)
analyses using `psborrow2`.

# Before you start

The `psborrows2` package relies on [`Stan`](https://mc-stan.org/) for model fitting,
specifically via the [`CmdStan`](https://mc-stan.org/users/interfaces/cmdstan.html)
and [`cmdstanr`](https://mc-stan.org/cmdstanr/) tools.

If you haven't used `CmdStan` before you'll need to install the R package
and the external program. More information can be found in the
[`cmdstanr` vignette](https://mc-stan.org/cmdstanr/articles/cmdstanr.html).

The short version is:
```{r install-cmdstan-2, eval=FALSE}
# Install the cmdstanr package
install.packages("cmdstanr", repos = c("https://mc-stan.org/r-packages/", getOption("repos")))
library(cmdstanr)

# Install the external CmdStan program
check_cmdstan_toolchain()
install_cmdstan(cores = 2)
```

Now you're ready to start with `psborrow2`.
```{r dependencies, message = FALSE}
library(psborrow2)
```

# Creating an analysis object {.tabset}
For a BDB analysis in `psborrow2`, we need to create an object
of class `Analysis` which contains all the information needed to build a model
and compile an MCMC sampler using Stan. To create an `Analysis` object,
we will call the function `create_analysis_obj()`. Let's look at the four
required arguments to this function and evaluate them one-at-a-time.

```
create_analysis_obj(
  data_matrix,
  outcome,
  borrowing,
  treatment
)
```

## 1. `data_matrix` {.tabset}

### Required elements

`data_matrix` is where we input the one-row-per-patient `numeric` matrix for our
analysis. The column names of the matrix are not fixed, so the names of columns
will be specified in the outcome, treatment, and borrowing sections.

There are two columns required for all analyses:

* A flag denoting receipt of the experimental intervention (`1`) or not (`0`)
* A flag denoting whether the patient was part of the external data source
(`1`) or the internal trial (`0`)

#### Time-to-event

If the outcome is time-to-event, then two additional columns are needed:

* The duration of follow-up for each patient
* A flag denoting whether the patient was censored (`1`) or not (`0`)

#### Binary endpoints

If the outcome is binary, one additional column is needed:

* A flag denoting whether a patient had the event of interest (`1`) or not (`0`)

#### Covariates

Covariates may also be included in BDB analyses. These should be included
in the data matrix if the plan is to adjust for them.

***Note*** Only `numeric` matrices are supported. See [Example data] for creating such a matrix
from a `data.frame`.

***Note*** No missing data is currently allowed, all values must be non-missing.

### Example data

We will be using an example dataset stored in `psborrow2`
(`example_matrix`). If you are starting from a data frame or tibble,
you can easily create a suitable matrix with the `psborrow2` helper function
`create_data_matrix()`.

#### Creating a data matrix with `create_data_matrix()`
```{r, eval = require(survival)}
# Start with data.frame
diabetic_df <- survival::diabetic

# For demonstration purposes, let some patients be external controls
diabetic_df$external <- ifelse(diabetic_df$trt == 0 & diabetic_df$id > 1000, 1, 0)

# Create the censor flag
diabetic_df$cens <- ifelse(diabetic_df$status == 0, 1, 0)

diabetes_matrix <- create_data_matrix(
  diabetic_df,
  outcome = c("time", "cens"),
  trt_flag_col = "trt",
  ext_flag_col = "external",
  covariates = ~ age + laser + risk
)

head(diabetes_matrix)
```

#### `psborrow2` example matrix
Let's look at the first few rows of the example matrix:

```{r example-matrix}
head(example_matrix)
```

The column definitions are below:

* `ext`, 0/1, flag for external controls
* `trt`, 0/1, flag for treatment arm
* `cov1`, 0/1, a baseline covariate
* `cov2`, 0/1, a baseline covariate
* `time`, positive numeric, survival time
* `cnsr`, 0/1, censoring indicator
* `resp`, 0/1, indicator for binary response outcome

## 2. `outcome`

`psborrow2` currently supports three outcomes:

* Time-to-event with exponential distribution (constant hazard), created with
`exp_surv_dist()`
* Time-to-event with Weibull distribution and proportional hazards
parametrization, created with `weib_ph_surv_dist()`
* Binary endpoints with a Bernoulli distribution and using logistic regression,
created with `logistic_bin_outcome()`

After we select which outcome and distribution we want,
we need to specify a prior distribution for the baseline event rate,
`baseline_prior`. In this case, `baseline_prior` is
a log hazard rate (for binary endpoints this represents the log
odds instead). Let's assume we have no prior knowledge on this event rate,
so we'll specify an uninformative prior: `normal_prior(0, 1000)`.

For our example, let's conduct a time-to-event analysis using the exponential
distribution.

```{r exp-surv-dist}
outcome <- exp_surv_dist(
  time_var = "time",
  cens_var = "cnsr",
  baseline_prior = normal_prior(0, 1000)
)
outcome
```

## 3. `borrowing`

`psborrow2` supports three different borrowing methods which are specified in
the `method` argument of `borrowing_details()` (see the code example below
how to call it):

* "No borrowing": This is the internal trial comparison without any external
data
* "Full borrowing": This is pooling of the external and internal control arms
* "BDB": Bayesian dynamic borrowing

The column name for the external control column flag in our matrix
is also required and passed to `ext_flag_col`.

Finally, for BDB only, the hyperprior distribution on the commensurability
parameter must be specified. This hyperprior determines
(along with the comparability of the outcomes
between internal and external controls) how much borrowing of the external
control group will be performed. Example hyperpriors include largely
uninformative inverse gamma distributions
e.g., `gamma_prior(alpha = .001, beta = .001)` as well as more
informative distributions e.g., `gamma_prior(alpha = 1, beta = .001)`,
though any distribution on the positive real line can be used. Distributions
with more density at higher values (i.e., higher precision)
will lead to more borrowing. We'll choose an uninformative gamma prior
in this example.

***Note***: Prior distributions are outlined in greater detail in a separate
vignette, see `vignette('prior_distributions', package = 'psborrow2')`.

```{r borrowing-details}
borrowing <- borrowing_details(
  method = "BDB",
  ext_flag_col = "ext",
  tau_prior = gamma_prior(0.001, 0.001)
)
borrowing
```

## 3. `treatment`

Finally, treatment details are outlined in `treatment_details()`. Here, we first
specify the column for the treatment flag in `trt_flag_col`. In addition, we
need to specify the prior on the effect estimate, `trt_prior`. We'll use
another uninformative normal distribution for the prior on the treatment effect:

```{r treatment-details}
treatment <- treatment_details(
  trt_flag_col = "trt",
  trt_prior = normal_prior(0, 1000)
)
treatment
```

## Application

Now that we have thought through each of the inputs to `create_analysis_obj()`,
let's create an analysis object:

```{r anls-obj}
anls_obj <- create_analysis_obj(
  data_matrix = example_matrix,
  outcome = outcome,
  borrowing = borrowing,
  treatment = treatment
)
```

The Stan model compiled successfully and informed us that we are
ready to begin sampling.

# Sampling from an analysis object

We can take draws from the posterior
distribution using the function `mcmc_sample()`. This function takes as input
our `Analysis` object and any arguments (other than the `data` argument)
that are passed to
[`CmdStanModel` objects](https://mc-stan.org/cmdstanr/reference/model-method-sample.html).
Note that running this may take a few minutes.

```{r res}
results <- mcmc_sample(anls_obj,
  iter_warmup = 2000,
  iter_sampling = 50000,
  chains = 4,
  seed = 112233
)
```

# Summarizing results{.tabset}

As a `CmdStanMCMC` object, `results` has several methods which are outlined on
the
[`cmdstanr` website](https://mc-stan.org/cmdstanr/reference/CmdStanModel.html).
For instance, we can see a see a summary of the posterior distribution samples
with `results$summary()`:

```{r res-summary}
results$summary()
```

The summary includes information for several parameter estimates
from our BDB model. Because it may not be immediately clear what the parameters
from the Stan model refer to, `psborrow2` has a function which returns a
variable dictionary from the analysis object to help interpret these parameters:

```{r data-dict}
variable_dictionary(anls_obj)
```

We can also capture all of the draws by calling `results$draws()`, which
returns an object of class `draws`. `draws` objects are
common in many MCMC sampling software packages and allow us to leverage
packages such as `posterior` and `bayesplot`.

```{r draws}
draws <- results$draws()
print(draws)
```

`psborrow2` also has a function to rename variables in `draws`
objects to be more interpretable, `rename_draws_covariates()`. This function
uses the `variable_dictionary` labels. Let's use it here to make the results
easier to interpret:

```{r rename-draws}
draws <- rename_draws_covariates(draws, anls_obj)
summary(draws)
```

## Using `bayesplot`

With `draws` objects and the `bayesplot` package, we can create many useful visual summary
plots. We can visualize the marginal posterior distribution of a
variable we are interested in by plotting histograms of the draws with the
function `mcmc_hist()`. Let's do that for the Hazard ratio for the treatment
effect and for our commensurability parameter, tau.

```{r mcmc-hist, message = FALSE}
library(bayesplot)
mcmc_hist(draws, c("treatment HR", "commensurability parameter"))
```

We can see other plots that help us understand and diagnose problems with
the MCMC sampler, such as trace and rank plots:

```{r mcmc-trace, fig.width=6, fig.height=3.6, dpi = 100}
color_scheme_set("mix-blue-pink")

mcmc_trace(
  draws[1:5000, 1:2, ], # Using a subset of draws only
  pars = c("treatment HR", "commensurability parameter"),
  n_warmup = 1000
)
```

Many other functions are outlined in the
[`bayesplot` vignettes](https://mc-stan.org/bayesplot/).

## Using `posterior`

`draws` objects are also supported by the `posterior` package,
which provides many other tools for analyzing MCMC draw data. For instance, we
can use the `summarize_draws()` function to derive 80% credible intervals
for all parameters:

```{r posterior-summ, message = FALSE}
library(posterior)
summarize_draws(draws, ~ quantile(.x, probs = c(0.1, 0.9)))
```

Another useful application of the `posterior` package is the evaluation of the Monte Carlo
standard error for quantiles of a variable of interest:

```{r mcse-mean}
vm <- extract_variable_matrix(draws, "treatment HR")
mcse_quantile(x = vm, probs = c(0.1, 0.5, 0.9))
```

`posterior` contains many other helpful functions, as outlined in their
[vignettes](https://cran.r-project.org/web/packages/posterior/vignettes/posterior.html).
