---
title: "psborrow2 Overview"
subtitle: "Matt Secrest, Isaac Gravestock"
output: 
  rmarkdown::ioslides_presentation:
    widescreen: true
vignette: >
  %\VignetteIndexEntry{5. Walkthrough overview slides}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introducing `psborrow2`

## Presentation flow

- Bayesian Dynamic Borrowing
- History of `psborrow` and `psborrow2`
- Overview of package structure
- Demonstration of a single dataset analysis
- Demonstration of a simulation study analysis

## Acknowledgements

`psborrow2` is the culmination of work from many individuals within 
industry, academia, and the FDA. The authors would like to thank,
among others, the below:

- Daniel Sabanes Bove
- Isaac Gravestock
- Craig Gower-Page
- Yichen Lu
- Herb Pang
- Matthew Secrest
- Jiawen Zhu

# Bayesian Dynamic Borrowing

## Backround

### <u><b>Challenge</b></u>
Difficult to recruit patients in the control arm

- Rare disease
- Pediatric condition
- Oncology

### <u><b>Opportunity</b></u>
Abundant data available on control arm

- Reduce randomization to control arm
- Improve statistical power of randomized trial

# In the evening

## Dinner

- Eat spaghetti
- Drink wine

---

```{r, cars, fig.cap="A scatterplot.", echo=FALSE}
plot(cars)
```

## Going to sleep

- Get in bed
- Count sheep
