---
title: "Incorporating propensity scores analysis in psborrow2"
author: "Isaac Gravestock"
output:rmarkdown::html_vignette:
vignette: >
  %\VignetteIndexEntry{5. Incorporating propensity scores analysis in psborrow2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
link-citations: yes
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 4,
  fig.align = "center"
)
```

```{r load, message=FALSE}
library(psborrow2)
```

Propensity scores (PS) methods offer various ways to adjust analyses for differences in groups of patients.

@austin2013 discusses various approaches for using PS with survival analyses to obtain effect measures similar to
randomized controlled trials.

@wang2021 discuss using PS for IPTW, matching and stratification in combination with a Bayesian analysis. These methods
allow for the separation of the design and analysis into two stages, which may be attractive in a regulatory setting.

Another approach is the direct inclusion of the PS as a covariate in the outcome model.

## Propensity Score Calculation

Propensity scores are the probabilities that an observation belongs to one group or another, in this setting the
internal or external patients. The PS models are fitted based on baseline covariates that could be confounders between
outcome and being in the internal/external group.

The simplest approach for calculating PS is a logistic generalized linear model. First, let's load some demo data from
`psborrow2`.

```{r data}
my_data <- example_surv
head(my_data)
```

Let's now fit a logistic generalized linear model and add the PS as an additional covariate.

```{r glm}
ps_model <- glm(ext ~ cov1 + cov2 + cov3, data = my_data, family = binomial)
ps <- predict(ps_model, type = "response")
my_data$ps <- ps
head(my_data)
```

The calculated PS can be included as a covariate in the analysis object, for the "covariate adjustment" PS approach:

```{r analysis_obj}
data_mat <- create_data_matrix(
  my_data,
  covariates = c("cov1", "cov2", "cov3", "ps"),
  ext_flag_col = "ext",
  outcome = c("eventtime", "censor"),
  trt_flag_col = "trt"
)

analysis_object <- create_analysis_obj(
  data_matrix = data_mat,
  covariates = add_covariates(c("cov1", "cov2", "cov3", "ps"),
    priors = normal_prior(0, 100)
  ),
  borrowing = borrowing_details("BDB",
    ext_flag_col = "ext",
    tau_prior = gamma_prior(0.1, 0.1)
  ),
  treatment = treatment_details("trt", normal_prior(0, 100)),
  outcome = weib_ph_surv_dist("eventtime", "censor",
    shape_prior = exponential_prior(1),
    baseline_prior = normal_prior(0, 1000)
  )
)
```

## Other PS Analyses

For more complex PS analyses, we can use one of the dedicated R packages.

### PSweight

The [PSweight](https://cran.r-project.org/package=PSweight) package can calculate PS using other model fitting
approaches than the simple glm described above. It also provides some summary statistics and plots.
For example we can use a generalized boosted regression model (gbm):

```{r psweight}
library(PSweight)
my_data2 <- example_surv
psw_model <- SumStat(ext ~ cov1 + cov2 + cov3, data = my_data2, method = "gbm")
summary(psw_model)
```

A plot of the distributions shows there is good overlap between the scores between both groups.

```{r psweight_plot, fig.show="hold", out.width="50%", results='hide', message = FALSE}
plot(psw_model, type = "density")
```

The PS values can be copied into the data set and the analysis object can be constructed as before.

```{r covs, results='hide'}
my_data2$ps <- psw_model$propensity[, 2]
```

This package can also trim the data.

```{r trim}
psw_trimmed <- PStrim(ext ~ cov1 + cov2 + cov3, data = my_data2, delta = 0.1)
psw_trimmed
```

Since there is a reasonable overlap between the data, only 5 patients were trimmed.

As before this can be used for in the analysis by creating a suitable data matrix.

```{r cov_mat}
create_data_matrix(
  psw_trimmed$data,
  covariates = c("cov1", "cov2", "cov3", "ps"),
  ext_flag_col = "ext",
  outcome = c("eventtime", "censor"),
  trt_flag_col = "trt"
)
```

### MatchIt

A variety of matching methods, including PS matching are implemented in the
[MatchIt](https://cran.r-project.org/package=MatchIt) package.

As described in the [Getting Started vignette](https://cran.r-project.org/web/packages/MatchIt/vignettes/MatchIt.html),
it can be useful to check the imbalance before matching.

```{r matchit}
library(MatchIt)
# No matching; constructing a pre-match matchit object
no_match <- matchit(trt ~ cov1 + cov2 + cov3,
  data = example_surv,
  method = NULL, distance = "glm"
)
summary(no_match)
```

Here we are matching treated to untreated to select the most comparable control group, regardless of whether they are
internal or external.

Now we can try a 1:1 nearest matching approach.

```{r matchit11}
match_11 <- matchit(trt ~ cov1 + cov2 + cov3,
  data = example_surv,
  method = "nearest", distance = "glm"
)
summary(match_11)
```

```{r matchit_jitter}
plot(match_11, type = "jitter", interactive = FALSE)
```

```{r matchit_qq}
plot(match_11,
  type = "qq", interactive = FALSE,
  which.xs = c("cov1", "cov2", "cov3")
)
```

Determining whether the balance after matching is appropriate is beyond the scope of this vignette. You can read more in
the [MatchIt Assessing Balance](https://cran.r-project.org/web/packages/MatchIt/vignettes/assessing-balance.html)
vignette. Another useful package is [cobalt](https://cran.r-project.org/package=cobalt), which provides tools for
assessing balance between groups after weighting or matching. It is compatible with many matching and weighting packaes.
See the [vignette](https://cran.r-project.org/web/packages/cobalt/vignettes/cobalt.html) for more details.

However, if you are happy with the results of the matching procedure, you can extract the data for use in psborrow2.

```{r trimmed_mat}
create_data_matrix(
  data = example_surv[match_11$weights == 1, ],
  covariates = c("cov1", "cov2", "cov3"),
  ext_flag_col = "ext",
  outcome = c("eventtime", "censor"),
  trt_flag_col = "trt"
)
```


# Weighted Analysis

The models also support fixed weights on the likelihood contributions from each observation. This
is equivalent to fixed power prior weights. This allows for the combination of models, such as an
IPTW + commensurate prior approach.

This also allows for power prior analyses:

```{r}
# fixed borrowing of external patients at alpha = 0.1
data_mat <- cbind(data_mat[, 1:8], w01 = ifelse(data_mat[, "ext"] == 1, 0.1, 1))


analysis_object_w01 <- create_analysis_obj(
  data_matrix = data_mat,
  covariates = add_covariates(c("cov1", "cov2"),
    priors = normal_prior(0, 100)
  ),
  borrowing = borrowing_details("Full borrowing",
    ext_flag_col = "ext",
  ),
  treatment = treatment_details("trt", normal_prior(0, 100)),
  outcome = exp_surv_dist("eventtime", "censor",
    baseline_prior = normal_prior(0, 1000),
    weight_var = "w01"
  )
)

analysis_object_w05 <- create_analysis_obj(
  data_matrix = cbind(data_mat[, 1:8], w01 = ifelse(data_mat[, "ext"] == 1, 0.5, 1)),
  covariates = add_covariates(c("cov1", "cov2"),
    priors = normal_prior(0, 100)
  ),
  borrowing = borrowing_details("Full borrowing",
    ext_flag_col = "ext",
  ),
  treatment = treatment_details("trt", normal_prior(0, 100)),
  outcome = exp_surv_dist("eventtime", "censor",
    baseline_prior = normal_prior(0, 1000),
    weight_var = "w01"
  )
)

analysis_object_w08 <- create_analysis_obj(
  data_matrix = cbind(data_mat[, 1:8], w01 = ifelse(data_mat[, "ext"] == 1, 0.8, 1)),
  covariates = add_covariates(c("cov1", "cov2"),
    priors = normal_prior(0, 100)
  ),
  borrowing = borrowing_details("Full borrowing",
    ext_flag_col = "ext",
  ),
  treatment = treatment_details("trt", normal_prior(0, 100)),
  outcome = exp_surv_dist("eventtime", "censor",
    baseline_prior = normal_prior(0, 1000),
    weight_var = "w01"
  )
)

analysis_object_w08 <- create_analysis_obj(
  data_matrix = cbind(data_mat[, 1:8], w01 = ifelse(data_mat[, "ext"] == 1, 0.8, 1)),
  covariates = add_covariates(c("cov1", "cov2"),
    priors = normal_prior(0, 100)
  ),
  borrowing = borrowing_details("Full borrowing",
    ext_flag_col = "ext",
  ),
  treatment = treatment_details("trt", normal_prior(0, 100)),
  outcome = exp_surv_dist("eventtime", "censor",
    baseline_prior = normal_prior(0, 1000),
    weight_var = "w01"
  )
)

analysis_object_tau2 <- create_analysis_obj(
  data_matrix = data_mat,
  covariates = add_covariates(c("cov1", "cov2"),
    priors = normal_prior(0, 100)
  ),
  borrowing = borrowing_details("BDB",
    ext_flag_col = "ext",
    tau_prior = uniform_prior(2, 2.00001)
  ),
  treatment = treatment_details("trt", normal_prior(0, 100)),
  outcome = exp_surv_dist("eventtime", "censor",
    baseline_prior = normal_prior(0, 1000)
  )
)


analysis_object_full <- create_analysis_obj(
  data_matrix = cbind(data_mat[, 1:8], w01 = ifelse(data_mat[, "ext"] == 1, 0.8, 1)),
  covariates = add_covariates(c("cov1", "cov2"),
    priors = normal_prior(0, 100)
  ),
  borrowing = borrowing_details("Full borrowing",
    ext_flag_col = "ext",
  ),
  treatment = treatment_details("trt", normal_prior(0, 100)),
  outcome = exp_surv_dist("eventtime", "censor",
    baseline_prior = normal_prior(0, 1000)
  )
)

analysis_object_no <- create_analysis_obj(
  data_matrix = cbind(data_mat[, 1:8], w01 = ifelse(data_mat[, "ext"] == 1, 0.8, 1)),
  covariates = add_covariates(c("cov1", "cov2"),
    priors = normal_prior(0, 100)
  ),
  borrowing = borrowing_details("No borrowing",
    ext_flag_col = "ext",
  ),
  treatment = treatment_details("trt", normal_prior(0, 100)),
  outcome = exp_surv_dist("eventtime", "censor",
    baseline_prior = normal_prior(0, 1000)
  )
)




pp_01 <- mcmc_sample(analysis_object_w01, verbose = TRUE)
pp_05 <- mcmc_sample(analysis_object_w05, verbose = TRUE)
pp_08 <- mcmc_sample(analysis_object_w08, verbose = TRUE)
pp_10 <- mcmc_sample(analysis_object_full, verbose = TRUE)
pp_00 <- mcmc_sample(analysis_object_no, verbose = TRUE)
tau_2 <- mcmc_sample(analysis_object_tau2, verbose = TRUE)

pp_00
pp_01
pp_05
pp_08
pp_10
tau_2
```


```{r}
library(survival)
exp_surv <- survival::survreg(
  Surv(eventtime, 1 - censor) ~ trt + cov1 + cov2,
  data = as.data.frame(data_mat),
  dist = "exponential"
)
summary(exp_surv)
```

An IPTW analysis

## References
