---
title: "Design for other borrowing models"
author: "Isaac Gravestock"
date: "2022-12-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Problem

Currently we try to build the stan model with the functions `make_model_string_data()`,
`make_model_string_parameters()`, etc. These functions insert code from the various
model parts, ie `Covariates`, `Outcome`, `Treatment`, etc. For the most part this works cleanly.
This approach becomes difficult in `make_model_string_model()`, which needs to integrate the
outcome model, the borrowing model, covariate adjustment and all priors. Currently this
has various cases hard coded, eg for full/no/BDB and for covariates/no covariates, but this doesn't
extensible and flexible enough for different models.

## Proposal

Let `make_model_string_model` be a generic with the signature taking the borrowing method
and the outcome model. Any future outcome or borrowing method can define those construction
functions as necessary. For some new outcome models, no additional method definition will
be required and it can just define that it uses an existing function.
Also borrowing methods will define for which outcome models they work.
This will also allow to explicitly show which models are not supported/implemented.

For the generic to work, we will also need to generalise `Borrowing` and define child classes
for each method.

Then our generic can be
```{r}
setGeneric("make_model_string_model", function(borrowing, outcome, analysis_obj) {
  standardGeneric("make_model_string_model")
})
```

The existing `make_model_string_model` can be carried over into this new model or broken into
smaller parts without any if/else constructions.


## Sketch

### Borrowing classes

```{r}
setClass(
  "Borrowing",
  slots = c(
    ext_flag_col = "character",
    data_stan_code = "character"
  )
)

# Dynamic
setClass("BDB", contains = "Borrowing", slots = c(tau_prior = "Prior"))
dynamic_borrowing <- function(tau_prior, ext_flag_col) {
  # check args
  new("BDB", ext_flag_col = ext_flag_col, tau_prior = tau_prior, data_stan_code = "matrix[N] Z;")
}

# Full Borrowing
setClass("Full_Borrowing", contains = "Borrowing")
full_borrowing <- function(ext_flag_col) {
  # check args
  new("Full_Borrowing", ext_flag_col = ext_flag_col, data_stan_code = "")
}

# No Borrowing
setClass("No_Borrowing", contains = "Borrowing")
no_orrowing <- function(ext_flag_col) {
  # check args
  new("No_Borrowing", ext_flag_col = ext_flag_col, data_stan_code = "")
}

# Commensurate All Betas
setClass("Commensurate", contains = "Borrowing", slots = c(tau_prior = "Prior"))
commensurate <- function(tau_prior, ext_flag_col) {
  # check args
  new("BDB", ext_flag_col = ext_flag_col, tau_prior = tau_prior, data_stan_code = "matrix[N] Z;")
}
```

### Model string method

```{r}
setMethod(
  "make_model_string_model",
  signature("BDB", "ANY", "Analysis"),
  function(borrowing, outcome, analysis_obj) {
    ### Treatment prior
    beta_trt_prior <- get_prior_string(analysis_obj@treatment@trt_prior)

    ### Linear predictor
    has_covariates <- !is.null(analysis_obj@covariates)

    if (has_covariates) {
      linear_predictor <- h_glue("
      lp = X * beta + Z * alpha + trt * beta_trt;
      elp = exp(lp) ;")
    } else if (!has_covariates) {
      linear_predictor <- h_glue("
      lp = Z * alpha + trt * beta_trt;
      elp = exp(lp) ;")
    }

    ### Add priors for relevant parameters
    if (NROW(analysis_obj@outcome@param_priors) > 0) {
      names <- names(analysis_obj@outcome@param_priors)
      values <- get_prior_string(analysis_obj@outcome@param_priors)
      outcome_prior <- h_glue("{{names}} ~ {{values}} ;", collapse = TRUE)
    } else {
      outcome_prior <- ""
    }

    ### Add priors on betas
    if (has_covariates) {
      i <- seq_along(analysis_obj@covariates@covariates)
      value <- get_prior_string(analysis_obj@covariates@priors)
      index <- if (test_named(value)) get_vars(analysis_obj@covariates) else rep(1, length(i))
      covariate_prior <- h_glue("beta[{{i}}] ~ {{value[index]}} ;", collapse = TRUE)
    } else {
      covariate_prior <- ""
    }

    tau_prior <- get_prior_string(analysis_obj@borrowing@tau_prior)
    alpha_2_prior <- get_prior_string(analysis_obj@outcome@baseline_prior)

    borrowing_string <- h_glue("
      tau ~ {{tau_prior}} ;
      real sigma;
      sigma = 1 / tau;
      alpha[2] ~ {{alpha_2_prior}} ;
      alpha[1] ~ normal(alpha[2], sqrt(sigma)) ;")

    ### Add in likelihood function
    likelihood_string <- h_glue("{{analysis_obj@outcome@likelihood_stan_code}}")

    h_glue("
  model {
    vector[N] lp;
    vector[N] elp;
    beta_trt ~ {{beta_trt_prior}};
    {{linear_predictor}}
    {{outcome_prior}}
    {{covariate_prior}}
    {{borrowing_string}}
    {{likelihood_string}}
  }")
  }
)



full_no_borrowing <- function(borrowing, outcome, analysis_obj) {
  ### Treatment prior
  beta_trt_prior <- get_prior_string(analysis_obj@treatment@trt_prior)

  ### Linear predictor
  has_covariates <- !is.null(analysis_obj@covariates)
  if (has_covariates) {
    linear_predictor <- h_glue("
      lp = alpha + X * beta + trt * beta_trt ;
      elp = exp(lp) ;")
  } else if (!has_covariates) {
    linear_predictor <- h_glue("
      lp = alpha + trt * beta_trt ;
      elp = exp(lp);")
  }

  ### Add priors for relevant parameters
  if (NROW(analysis_obj@outcome@param_priors) > 0) {
    names <- names(analysis_obj@outcome@param_priors)
    values <- get_prior_string(analysis_obj@outcome@param_priors)
    outcome_prior <- h_glue("{{names}} ~ {{values}} ;", collapse = TRUE)
  } else {
    outcome_prior <- ""
  }

  ### Add priors on betas
  if (has_covariates) {
    i <- seq_along(analysis_obj@covariates@covariates)
    value <- get_prior_string(analysis_obj@covariates@priors)
    index <- if (test_named(value)) get_vars(analysis_obj@covariates) else rep(1, length(i))
    covariate_prior <- h_glue("beta[{{i}}] ~ {{value[index]}} ;", collapse = TRUE)
  } else {
    covariate_prior <- ""
  }

  alpha_prior <- get_prior_string(analysis_obj@outcome@baseline_prior)
  borrowing_string <- h_glue("alpha ~ {{alpha_prior}} ;")

  ### Add in likelihood function
  likelihood_string <- h_glue("{{analysis_obj@outcome@likelihood_stan_code}}")

  h_glue("
  model {
    vector[N] lp;
    vector[N] elp;
    beta_trt ~ {{beta_trt_prior}};
    {{linear_predictor}}
    {{outcome_prior}}
    {{covariate_prior}}
    {{borrowing_string}}
    {{likelihood_string}}
  }")
}

setMethod("make_model_string_model", signature("Full_Borrowing", "ANY", "Analysis"), full_no_borrowing)
setMethod("make_model_string_model", signature("No_Borrowing", "ANY", "Analysis"), full_no_borrowing)
```


So far no new functionality. Looking at the above code we can probably factor our some helper code for priors.

The benefit comes now that we can define a method for `"BDB'` for piecewise exponential without complicating the other
code.

```{r}
setMethod(
  "make_model_string_model",
  signature("BDB", "PiecewiseExponentialSurvDist", "Analysis"),
  function(borrowing, outcome, analysis_obj) {
    ### Treatment prior
    beta_trt_prior <- get_prior_string(analysis_obj@treatment@trt_prior)

    ### Linear predictor
    has_covariates <- !is.null(analysis_obj@covariates)

    linear_predictor <- if (has_covariates) {
      h_glue("lp = Z*alpha_ext' + (1-Z)*alpha' + rep_matrix(trt * beta_trt + X * beta, M);")
    } else if (!has_covariates) {
      h_glue("lp = Z*alpha_ext' + (1-Z)*alpha' + rep_matrix(trt * beta_trt, M);")
    }

    ### Add priors for relevant parameters
    if (NROW(analysis_obj@outcome@param_priors) > 0) {
      names <- names(analysis_obj@outcome@param_priors)
      values <- get_prior_string(analysis_obj@outcome@param_priors)
      outcome_prior <- h_glue("{{names}} ~ {{values}} ;", collapse = TRUE)
    } else {
      outcome_prior <- ""
    }

    ### Add priors on betas
    if (has_covariates) {
      i <- seq_along(analysis_obj@covariates@covariates)
      value <- get_prior_string(analysis_obj@covariates@priors)
      index <- if (test_named(value)) get_vars(analysis_obj@covariates) else rep(1, length(i))
      covariate_prior <- h_glue("beta[{{i}}] ~ {{value[index]}} ;", collapse = TRUE)
    } else {
      covariate_prior <- ""
    }

    tau_prior <- get_prior_string(analysis_obj@borrowing@tau_prior)
    alpha_2_prior <- get_prior_string(analysis_obj@outcome@baseline_prior)

    borrowing_string <- h_glue("
      tau ~ {{tau_prior}} ;
      real sigma;
      sigma = 1 / tau;
      alpha_ext ~ {{alpha_2_prior}};
      alpha ~ normal(alpha_ext, sqrt(sigma)) ;
      ")

    ### Add in likelihood function
    likelihood_string <- "target += sum((lp .* D) - (exp(lp) .* T));"

    h_glue("
  model {
    matrix[N,M] lp;
    vector[N] elp;
    beta_trt ~ {{beta_trt_prior}};
    {{linear_predictor}}
    {{outcome_prior}}
    {{covariate_prior}}
    {{borrowing_string}}
    {{likelihood_string}}
  }")
  }
)


setMethod(
  "make_model_string_model",
  signature("No_Borrowing", "PiecewiseExponentialSurvDist", "Analysis"),
  function(borrowing, outcome, analysis_obj) {
    ### Treatment prior
    beta_trt_prior <- get_prior_string(analysis_obj@treatment@trt_prior)

    ### Linear predictor
    has_covariates <- !is.null(analysis_obj@covariates)

    linear_predictor <- if (has_covariates) {
      h_glue("lp = alpha' + rep_matrix(trt * beta_trt + X * beta, M);")
    } else if (!has_covariates) {
      h_glue("lp = alpha' + rep_matrix(trt * beta_trt, M);")
    }

    ### Add priors for relevant parameters
    if (NROW(analysis_obj@outcome@param_priors) > 0) {
      names <- names(analysis_obj@outcome@param_priors)
      values <- get_prior_string(analysis_obj@outcome@param_priors)
      outcome_prior <- h_glue("{{names}} ~ {{values}} ;", collapse = TRUE)
    } else {
      outcome_prior <- ""
    }

    ### Add priors on betas
    if (has_covariates) {
      i <- seq_along(analysis_obj@covariates@covariates)
      value <- get_prior_string(analysis_obj@covariates@priors)
      index <- if (test_named(value)) get_vars(analysis_obj@covariates) else rep(1, length(i))
      covariate_prior <- h_glue("beta[{{i}}] ~ {{value[index]}} ;", collapse = TRUE)
    } else {
      covariate_prior <- ""
    }

    tau_prior <- get_prior_string(analysis_obj@borrowing@tau_prior)
    alpha_prior <- get_prior_string(analysis_obj@outcome@baseline_prior)
    borrowing_string <- h_glue("alpha ~ {{alpha_prior}};")

    ### Add in likelihood function
    likelihood_string <- "target += sum((lp .* D) - (exp(lp) .* T));"

    h_glue("
  model {
    matrix[N,M] lp;
    beta_trt ~ {{beta_trt_prior}};
    {{linear_predictor}}
    {{outcome_prior}}
    {{covariate_prior}}
    {{borrowing_string}}
    {{likelihood_string}}
  }")
  }
)
```


### Classes for Piecewise exponential survival

```{r}
.piecewise_exp_surv_dist <- setClass(
  "PiecewiseExponentialSurvDist",
  contains = "TimeToEvent",
  slots = c(
    start_times = "numeric",
    transformed_data_stan_code = "character"
  ),
  prototype = list(
    n_param = 0L,
    likelihood_stan_code = ""
  )
)

#' Piecewise Exponential survival distribution
#'
#' @param time_var character. Name of time variable column in model matrix
#' @param cens_var character. Name of the censorship variable flag in model matrix
#' @param weight_var character. Optional name of variable in model matrix for weighting the log likelihood.
#' @param baseline_prior `Prior`. Object of class `Prior` specifying prior distribution for the baseline outcome.
#' @param start_times numeric. A vector of times to define periods.
#'
#' @return Object of class [`PiecewiseExponentialSurvDist`][PiecewiseExponentialSurvDist-class].
#' @export
#' @family outcome models
#'
#' @examples
#' es <- piecewise_exp_surv_dist(
#'   time_var = "time",
#'   cens_var = "cens",
#'   baseline_prior = normal_prior(0, 1000),
#'   start_times = c(0, 10, 20)
#' )
piecewise_exp_surv_dist <- function(time_var, cens_var, start_times, baseline_prior, weight_var = "") {
  assert_string(time_var)
  assert_string(cens_var)
  assert_string(weight_var)
  assert_numeric(start_times, lower = 0, finite = TRUE, any.missing = FALSE, min.len = 1, sorted = TRUE)
  assert_class(baseline_prior, "Prior")
  has_weight <- isTRUE(weight_var != "")
  .piecewise_exp_surv_dist(
    time_var = time_var,
    cens_var = cens_var,
    baseline_prior = baseline_prior,
    weight_var = weight_var,
    likelihood_stan_code = "",
    data_stan_code = h_glue(
      "
      vector[N] time;
      vector[N] cens;
      int<lower = 1> M = {{length(start_times)}};
      vector[M] starts = [{{toString(start_times)}}]';
      {{weight}}",
      weight = if (has_weight) "vector[N] weight;" else ""
    ),
    function_stan_code = h_glue("
    vector make_durations(vector starts, vector time) {
      vector [rows(starts)] ends = append_row(tail(starts, rows(starts) - 1), max(time));
      return fdim(ends, starts);
    }"),
    param_stan_code = "vector[M] alpha;",
    transformed_data_stan_code = h_glue("
      vector[M] durations;
      matrix[N,M] T;
      matrix[N,M] D;

      durations = make_durations(starts, time);

      for(j in 1:M) {
        T[,j] = fmin(fdim(time, starts[j]), durations[j]);
      }
      for(j in 1:M) {
        for(i in 1:N) {
          D[i,j] = (starts[j] <= time[i] && time[i] < starts[j] + durations[j]) * (1 - cens[i]);
        }
      }")
  )
}
```
