---
title: "Simulate Data: Baseline"
author: "Isaac Gravestock"
date: "2023-03-31"
output: html_document
---

# 1. Baseline Design

To prepare for simulating survival data, we need to first create the baseline data for the cohort.

Requirements:

1. number of patients in the 3 arms
2. arbitrary number of correlated variables
3. transform to binary variables by cut-off value
4. set names
5. set means for internal and external arms

```{r, eval = FALSE}
# generate correlated variables
correlated_covs <- add_covariates(
  names = c("x1", "x2", "x3"),
  means_int = c(0, 0, 0),
  means_ext = means,
  covariance = covariance_matrix(diag = c(1, 1, 1), upper_tri = c(0, .2, .3))
) %>% transform(x1 = binary_tranform(cut_off = 0.4, cut_off_ext = 0.45))

# generate uncorrelated variables
indep_covs <- add_covariates(
  names = c("age", "sex"),
  means = c(60, 0.5),
  means_ext = means,
  covariance = covariance_matrix(diag = c(1, 1), upper_tri = c(0, 0))
)

baseline_obj <- sim_baseline(
  sample_size = c(trt_int = 100, ctrl_int = 50, ctrl_ext = 200),
  covariates = list(correlated_covs, indep_covs)
)
```

`sim_baseline` is enough to specify a cohort to generate survival times.
Implementation of the above functions can be a first phase, generating survival data will be in another
design doc

# Functions descriptions

```{r}
# Specify covariates to include in baseline
#
# names: covariates names
# means_int: means for internal trial patients
# means_ext: means for external patients
# covariance_int: covariance matrix for multivariate normal for internal
# covariance_ext: covariance matrix for multivariate normal for external
#
# Returns object capturing these parameters for later processing
add_covariates(names, means_int, means_ext, covariance_int, covariance_ext)

# Transform variables
#
# ... Arguments corresponding to the names in `add_covariates` taking functions
#  which takes a list of vectors of percentiles (int, ext), transforms them, and returns
#  a list of vectors of the same dimension.
#
#  Returns a modified covariates object
transform(...)

binary_tranform <- function(cut_off_int, cut_off_ext) {
  function(qlist) {
    list(int = qlist$int > cut_off_int, ext = qlist$ext > cut_off_ext)
  }
}


# Covariance matrix
#
# diag: variances to put in the main diagonal
# upper_tri: covariance terms to put in the upper and lower triangles. Given by column
covariance_matrix(diag, upper_tri)

# Simulate Baseline
#
# sample_size: named vector of sample sizes (maybe these should be directly parameters)
# covariates: a list of covariates objects
#
# Constructs the multivariate normal distribution by combining the means and covariance matrices blockwise.
# Generates a data frame using mvrnorm().
# Any variables to be transformed and calculated back to percentiles using marginal distribution and then
# transformed.
sim_baseline(sample_size, covariates)
```
