---
title: 'Simulate Data: Survival'
author: "Isaac Gravestock"
date: "2023-06-20"
output: html_document
---

Continuing from the design of baseline covariates, we consider how to implement survival data simulation.

# 2. Survival Design

For survival we need to specify in addition:

 - baseline hazard parameters
 - HR:
   - internal/external HR "drift"
   - covariate HR coefficients
   - treatment effect HR

```{r}
with_age_score <- create_baseline_object(
  n_trt_int = 100,
  n_ctrl_int = 50,
  n_ctrl_ext = 100,
  covariates = baseline_covariates(
    names = c("age", "score"),
    means_int = c(55, 5),
    means_ext = c(60, 5),
    covariance_int = covariance_matrix(c(5, 1))
  )
)

with_age_score@transformations <- list(
  score_high = binary_cutoff("score", int_cutoff = 0.7, ext_cutoff = 0.7),
  score_rounded = function(data) round(data$score),
  score_high_factor = function(data) factor(round(data$score), levels = 0:10)
)

weibull_surv <- event_dist(dist = "weibull", lambdas = 1 / 200, gammas = 1)

my_data_sim <- create_data_simulation(
  with_age_score,
  coefficients = c(age = 0.001, score_high = 1.5),
  event_dist = weibull_surv,
)

my_data <- generate(
  my_data_sim,
  treatment_effect = c(0, .5, 1),
  drift = c(0, .2, .4)
)

my_data$guide
head(my_data$data_list[[1]][[1]])
plot(
  x = my_data$data_list[[1]][[1]][, "age"],
  y = my_data$data_list[[1]][[1]][, "eventtime"],
  col = my_data$data_list[[1]][[1]][, "score_high"] + 1
)
```


```{r}
# each of these builds up the object
# nothing needs to generate any data
#
# implement generate() method for all of them
# generate methods are recursive,
# highest level would take n=N repeat from highest level
#


baseline_obj %>%
  design_matrx(~ sex + trt + ext) %>%
  linear_predictor(betas = c(sex = 0.3, trt = 1, ext = 1.1)) %>%
  simulate_survival(hazard_model = weibull_ph(shape = 1, scale = 0.9)) %>%
  survival_matrix() %>% # think of better name....
  generate(n = 100)

generate(baseline_obj()) %>%
  mutate(age2 = age^2) %>%
  design_matrix(~ sex + trt + ext)

baseline_design <- design_matrix(baseline_obj, ~ sex + I(age^2) + trt + ext, preprocess_fun = function(data) data)
baseline_design2 <- design_matrix(baseline_obj, ~., preprocess_fun = function(data) data$age2 <- data$age^2)
colnames(baseline_design)


baseline_predictor <- linear_predictor(
  design = baseline_design, # matrix
  betas = c(
    sex = .3,
    "I(age^2)" = .4,
    trt = 1,
    ext = 1.1
  )
)

simulate_survival(
  baseline_predictor,
  hazard_model = weibull_ph(shape = 0.76, lambda = 1) # is a function that takes vector of LPs and gives vector of times
)

survival_matrix(
  baseline_design,
  surv_times = simulate_survival(baseline_predictor, hazard_model = weibull_ph(shape = 0.76, lambda = 1)),
  censor_times = censor(),
  recruitment = recruitment()
)
```

Refer to psborrow for how these can be calculated and combined.

- recruitment rate
- cut off
- censoring
