---
title: "8. Simple Overview"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{8. Simple Overview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Purpose
The goal of this demo is to conduct a Bayesian Dynamic Borrowing
(BDB) analysis using the hierarchical commensurate prior on a dataset.

```{r, include = FALSE}
# Load packages ----
library(psborrow2)

# Survival analysis
library(survival)
library(ggsurvfit)
library(flexsurv)

# Additional tools for draws objects
library(bayesplot)
library(posterior)

# Comparing populations
library(table1)

```

```
# Load packages ----
library(psborrow2)

# Survival analysis
library(survival)
library(ggsurvfit)
library(flexsurv)

# Additional tools for draws objects
library(bayesplot)
library(posterior)

# Comparing populations
library(table1)
```


# Explore example data

## {psborrow2} contains an example matrix
```{r}
head(example_matrix)
```

```
?example_matrix # for more details
```


## Load as data.frame for some functions
```{r}
example_dataframe <- as.data.frame(example_matrix)
```


## Look at distribution of arms
```{r}
table(ext = example_matrix[, "ext"], trt = example_matrix[, "trt"])
```

## Naive internal comparisons

### Kaplan-Meier curves
```{r}
km_fit <- survfit(Surv(time = time, event = 1 - cnsr) ~ trt + ext,
  data = example_dataframe
)

ggsurvfit(km_fit)
```

The internal and external control arms look quite different...

### Cox model
```{r}
cox_fit <- coxph(Surv(time = time, event = 1 - cnsr) ~ trt,
  data = example_dataframe,
  subset = ext == 0
)

cox_fit
exp(confint(cox_fit)) # The internal HR is 0.90 (95% CI 0.61 - 1.32)
```

# Hybrid control analysis

Let's start by demonstrating the utility of BDB by trying to
borrow data from the external control arm which we know
experiences worse survival.

The end goal is to create an Analysis object with:
```
?create_analysis_obj
```

# A note on prior distributions
psborrow2 allows the user to specify priors with the following
functions:

```
?prior_bernoulli
?prior_beta
?prior_cauchy
?prior_exponential
?prior_gamma
?prior_normal
?prior_poisson
?prior_uniform
```

Prior distributions can be plotted with the plot() method
```{r}
plot(prior_normal(0, 1), xlim = c(-100, 100), ylim = c(0, 1))
plot(prior_normal(0, 10), xlim = c(-100, 100), ylim = c(0, 1))
plot(prior_normal(0, 10000), xlim = c(-100, 100), ylim = c(0, 1))
```

# Outcome objects

psborrow2 currently supports 4 outcomes:

```
?outcome_surv_weibull_ph # Weibull survival w/ proportional hazards
?outcome_surv_exponential # Exponential survival
?outcome_bin_logistic # Logistic binary outcome
?outocme_cont_normal # Normal continuous outcome
```

# Create an exponential survival distribution Outcome object
```{r}
exp_outcome <- outcome_surv_exponential(
  time_var = "time",
  cens_var = "cnsr",
  baseline_prior = prior_normal(0, 10000)
)
```

# Borrowing object

Borrowing objects are created with:
```
?borrowing_hierarchical_commensurate # Hierarchical commensurate borrowing
?borrowing_none # No borrowing
?borrowing_full # Full borrowing
```

```{r}
bdb_borrowing <- borrowing_hierarchical_commensurate(
  ext_flag_col = "ext",
  tau_prior = prior_gamma(0.001, 0.001)
)
```

# Treatment objects

Treatment objects are created with:
```
?treatment_details
```

```{r}
trt_details <- treatment_details(
  trt_flag_col = "trt",
  trt_prior = prior_normal(0, 10000)
)
```

# Analysis objects

Combine everything and create object of class Analysis:
```{r}
analysis_object <- create_analysis_obj(
  data_matrix = example_matrix,
  outcome = exp_outcome,
  borrowing = bdb_borrowing,
  treatment = trt_details
)

analysis_object
```

# MCMC sampling

Conduct MCMC sampling with:
```
?mcmc_sample
```

```{r, include = FALSE}
results <- mcmc_sample(
  x = analysis_object,
  iter_warmup = 1000,
  iter_sampling = 5000,
  chains = 1
)
```

```
results <- mcmc_sample(
  x = analysis_object,
  iter_warmup = 1000,
  iter_sampling = 5000,
  chains = 1
)
```

```{r}
class(results)

results

```

# Interpret results


Dictionary to interpret parameters:
```{r}
variable_dictionary(analysis_object)
```

Create a draws object
```{r}
draws <- results$draws()
```

Rename draws object parameters
```{r}
draws <- rename_draws_covariates(draws, analysis_object)
```


Get 95% credible intervals with posterior package
```{r}
posterior::summarize_draws(draws, ~ quantile(.x, probs = c(0.025, 0.50, 0.975)))
```

Look at histogram of draws with bayesplot package
```{r}
bayesplot::mcmc_hist(draws, c("treatment HR"))
```

Our model does not borrow much from the external arm!
This is the ***desired outcome*** given how different the control arms were.

# Control arm imbalances

Check balance between arms
```{r}
table1(
  ~ cov1 + cov2 + cov3 + cov4 |
    factor(ext, labels = c("Internal RCT", "External data")) +
      factor(trt, labels = c("Not treated", "Treated")),
  data = example_dataframe
)
```

Because the imbalance may be conditional on observed covariates,
let's adjust for propensity scores in our analysis

Create a propensity score model
```{r}
ps_model <- glm(ext ~ cov1 + cov2 + cov3 + cov4,
  data = example_dataframe,
  family = binomial
)
ps <- predict(ps_model, type = "response")
example_dataframe$ps <- ps
example_dataframe$ps_cat_ <- cut(
  example_dataframe$ps,
  breaks = 5,
  include.lowest = TRUE
)
levels(example_dataframe$ps_cat_) <- c(
  "ref", "low",
  "low_med", "high_med", "high"
)
```


Convert the data back to a matrix with dummy variables for `ps_cat_` levels
```{r}
example_matrix_ps <- create_data_matrix(
  example_dataframe,
  outcome = c("time", "cnsr"),
  trt_flag_col = "trt",
  ext_flag_col = "ext",
  covariates = ~ps_cat_
)
```

Propensity score analysis without borrowing
```{r, include = FALSE}
anls_ps_no_borrow <- create_analysis_obj(
  data_matrix = example_matrix_ps,
  covariates = add_covariates(
    c("ps_cat_low", "ps_cat_low_med", "ps_cat_high_med", "ps_cat_high"),
    prior_normal(0, 10000)
  ),
  outcome = outcome_surv_exponential("time", "cnsr", prior_normal(0, 10000)),
  borrowing = borrowing_none("ext"),
  treatment = treatment_details("trt", prior_normal(0, 10000))
)

res_ps_no_borrow <- mcmc_sample(
  x = anls_ps_no_borrow,
  iter_warmup = 1000,
  iter_sampling = 5000,
  chains = 1
)
```

```
anls_ps_no_borrow <- create_analysis_obj(
  data_matrix = example_matrix_ps,
  covariates = add_covariates(
    c("ps_cat_low", "ps_cat_low_med", "ps_cat_high_med", "ps_cat_high"),
    prior_normal(0, 10000)
  ),
  outcome = outcome_surv_exponential("time", "cnsr", prior_normal(0, 10000)),
  borrowing = borrowing_none("ext"),
  treatment = treatment_details("trt", prior_normal(0, 10000))
)

res_ps_no_borrow <- mcmc_sample(
  x = anls_ps_no_borrow,
  iter_warmup = 1000,
  iter_sampling = 5000,
  chains = 1
)
```

```{r}
draws_ps_no_borrow <- rename_draws_covariates(
  res_ps_no_borrow$draws(),
  anls_ps_no_borrow
)

summarize_draws(draws_ps_no_borrow, ~ quantile(.x, probs = c(0.025, 0.50, 0.975)))
```


# Propensity score analysis with BDB

```{r, include = FALSE}
anls_ps_bdb <- create_analysis_obj(
  data_matrix = example_matrix_ps,
  covariates = add_covariates(
    c("ps_cat_low", "ps_cat_low_med", "ps_cat_high_med", "ps_cat_high"),
    prior_normal(0, 10000)
  ),
  outcome = outcome_surv_exponential("time", "cnsr", prior_normal(0, 10000)),
  borrowing = borrowing_hierarchical_commensurate("ext", prior_gamma(0.001, 0.001)),
  treatment = treatment_details("trt", prior_normal(0, 10000))
)

res_ps_bdb <- mcmc_sample(
  x = anls_ps_bdb,
  iter_warmup = 1000,
  iter_sampling = 5000,
  chains = 1
)

```

```
anls_ps_bdb <- create_analysis_obj(
  data_matrix = example_matrix_ps,
  covariates = add_covariates(
    c("ps_cat_low", "ps_cat_low_med", "ps_cat_high_med", "ps_cat_high"),
    prior_normal(0, 10000)
  ),
  outcome = outcome_surv_exponential("time", "cnsr", prior_normal(0, 10000)),
  borrowing = borrowing_hierarchical_commensurate("ext", prior_gamma(0.001, 0.001)),
  treatment = treatment_details("trt", prior_normal(0, 10000))
)

res_ps_bdb <- mcmc_sample(
  x = anls_ps_bdb,
  iter_warmup = 1000,
  iter_sampling = 5000,
  chains = 1
)
```

```{r}
draws_ps_bdb <- rename_draws_covariates(
  res_ps_bdb$draws(),
  anls_ps_bdb
)

summarize_draws(draws_ps_bdb, ~ quantile(.x, probs = c(0.025, 0.50, 0.975)))
```

It looks like PS + BDB allowed us to most accurately recover the
true hazard ratio of 0.70.
