---
title: "7. Comparison of Fixed Weights"
author: "Matt Secrest and Isaac Gravestock"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{7. Comparison of Fixed Weights}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

```{r knitr, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  fig.align = "center",
  fig.width = 5,
  fig.height = 3,
  dpi = 120,
  comment = "#"
)
options(rmarkdown.html_vignette.check_title = FALSE)
```

# Introduction

In a `psborrow2` analysis it is possible to specify fixed weights for an observation's log-likelihood contribution.
This is similar to a weighted regression or a fixed power prior parameter.

This vignette will show how weights can be specified and compare regression model results with other packages.
We will compare a `glm` model with weights, a weighted likelihood in Stan with `psborrow2`, and
`BayesPPD::glm.fixed.a0` for generalized linear models with fixed `a0` (power prior parameter).

```{r packages, results = "hide", message = FALSE}
library(psborrow2)
library(BayesPPD)
library(ggplot2)
```

# Logistic regression

We fit logistic regression models with the external control arm having weights (or power parameters)
equal to 0, 0.25, 0.5, 0.75, 1. The internal treated and control patients have weight = 1.
The model has a treatment indicator and two covariates, `resp ~ trt + cov1 + cov2`.

### glm

```{r glm_logistic, echo = FALSE, message = FALSE}
logistic_glm_reslist <- lapply(c(0, 0.25, 0.5, 0.75, 1), function(w) {
  logistic_glm <- glm(
    resp ~ trt + cov1 + cov2,
    data = as.data.frame(example_matrix),
    family = binomial,
    weights = ifelse(example_matrix[, "ext"] == 1, w, 1)
  )
  glm_summary <- summary(logistic_glm)$coef
  ci <- confint(logistic_glm)
  data.frame(
    fitter = "glm",
    borrowing = w,
    variable = c("(Intercept)", "trt", "cov1", "cov2"),
    estimate = glm_summary[, "Estimate"],
    lower = ci[, 1],
    upper = ci[, 2]
  )
})
```

```{r glm_logistic-no-eval, eval = FALSE}
logistic_glm_reslist <- lapply(c(0, 0.25, 0.5, 0.75, 1), function(w) {
  logistic_glm <- glm(
    resp ~ trt + cov1 + cov2,
    data = as.data.frame(example_matrix),
    family = binomial,
    weights = ifelse(example_matrix[, "ext"] == 1, w, 1)
  )
  glm_summary <- summary(logistic_glm)$coef
  ci <- confint(logistic_glm)
  data.frame(
    fitter = "glm",
    borrowing = w,
    variable = c("(Intercept)", "trt", "cov1", "cov2"),
    estimate = glm_summary[, "Estimate"],
    lower = ci[, 1],
    upper = ci[, 2]
  )
})
```

### BayesPPD

```{r bayesppd_logistic, echo = FALSE, message = FALSE}
logistic_ppd_reslist <- lapply(c(0, 0.25, 0.5, 0.75, 1), function(w) {
  logistic_ppd <- glm.fixed.a0(
    data.type = "Bernoulli",
    data.link = "Logistic",
    y = example_matrix[example_matrix[, "ext"] == 0, ][, "resp"],
    x = example_matrix[example_matrix[, "ext"] == 0, ][, c("trt", "cov1", "cov2")],
    historical = list(list(
      y0 = example_matrix[example_matrix[, "ext"] == 1, ][, "resp"],
      x0 = example_matrix[example_matrix[, "ext"] == 1, ][, c("cov1", "cov2")],
      a0 = w
    )),
    lower.limits = rep(-100, 5),
    upper.limits = rep(100, 5),
    slice.widths = rep(1, 5),
    nMC = 10000,
    nBI = 1000
  )[[1]]
  ci <- apply(logistic_ppd, 2, quantile, probs = c(0.025, 0.975))
  data.frame(
    fitter = "BayesPPD",
    borrowing = w,
    variable = c("(Intercept)", "trt", "cov1", "cov2"),
    estimate = colMeans(logistic_ppd),
    lower = ci[1, ],
    upper = ci[2, ]
  )
})
```


```{r bayesppd_logistic-no-eval, eval = FALSE}
logistic_ppd_reslist <- lapply(c(0, 0.25, 0.5, 0.75, 1), function(w) {
  logistic_ppd <- glm.fixed.a0(
    data.type = "Bernoulli",
    data.link = "Logistic",
    y = example_matrix[example_matrix[, "ext"] == 0, ][, "resp"],
    x = example_matrix[example_matrix[, "ext"] == 0, ][, c("trt", "cov1", "cov2")],
    historical = list(list(
      y0 = example_matrix[example_matrix[, "ext"] == 1, ][, "resp"],
      x0 = example_matrix[example_matrix[, "ext"] == 1, ][, c("cov1", "cov2")],
      a0 = w
    )),
    lower.limits = rep(-100, 5),
    upper.limits = rep(100, 5),
    slice.widths = rep(1, 5),
    nMC = 10000,
    nBI = 1000
  )[[1]]
  ci <- apply(logistic_ppd, 2, quantile, probs = c(0.025, 0.975))
  data.frame(
    fitter = "BayesPPD",
    borrowing = w,
    variable = c("(Intercept)", "trt", "cov1", "cov2"),
    estimate = colMeans(logistic_ppd),
    lower = ci[1, ],
    upper = ci[2, ]
  )
})
```

### psborrow2

```{r psb2_logistic, echo = FALSE, message = FALSE}
logistic_psb_reslist <- lapply(c(0, 0.25, 0.5, 0.75, 1), function(w) {
  logistic_psb2 <- create_analysis_obj(
    data_matrix = as.matrix(cbind(
      example_matrix,
      w = ifelse(example_matrix[, "ext"] == 1, w, 1)
    )),
    covariates = add_covariates(c("cov1", "cov2"),
      priors = normal_prior(0, 100)
    ),
    borrowing = borrowing_details("Full borrowing",
      ext_flag_col = "ext",
    ),
    treatment = treatment_details("trt", normal_prior(0, 100)),
    outcome = outcome_bin_logistic("resp",
      baseline_prior = normal_prior(0, 1000),
      weight_var = "w"
    )
  )

  mcmc_logistic_psb2 <- mcmc_sample(logistic_psb2, chains = 1, verbose = FALSE, seed = 123)
  mcmc_summary <- mcmc_logistic_psb2$summary(
    variables = c("alpha", "beta_trt", "beta[1]", "beta[2]"),
    mean,
    ~ quantile(.x, probs = c(0.025, 0.975))
  )

  data.frame(
    fitter = "psborrow2",
    borrowing = w,
    variable = c("(Intercept)", "trt", "cov1", "cov2"),
    estimate = mcmc_summary$mean,
    lower = mcmc_summary$`2.5%`,
    upper = mcmc_summary$`97.5%`
  )
})
```


```{r psb2_logistic-no-eval, eval = FALSE}
logistic_psb_reslist <- lapply(c(0, 0.25, 0.5, 0.75, 1), function(w) {
  logistic_psb2 <- create_analysis_obj(
    data_matrix = as.matrix(cbind(
      example_matrix,
      w = ifelse(example_matrix[, "ext"] == 1, w, 1)
    )),
    covariates = add_covariates(c("cov1", "cov2"),
      priors = normal_prior(0, 100)
    ),
    borrowing = borrowing_details("Full borrowing",
      ext_flag_col = "ext",
    ),
    treatment = treatment_details("trt", normal_prior(0, 100)),
    outcome = outcome_bin_logistic("resp",
      baseline_prior = normal_prior(0, 1000),
      weight_var = "w"
    )
  )

  mcmc_logistic_psb2 <- mcmc_sample(logistic_psb2, chains = 1, verbose = FALSE, seed = 123)
  mcmc_summary <- mcmc_logistic_psb2$summary(
    variables = c("alpha", "beta_trt", "beta[1]", "beta[2]"),
    mean,
    ~ quantile(.x, probs = c(0.025, 0.975))
  )

  data.frame(
    fitter = "psborrow2",
    borrowing = w,
    variable = c("(Intercept)", "trt", "cov1", "cov2"),
    estimate = mcmc_summary$mean,
    lower = mcmc_summary$`2.5%`,
    upper = mcmc_summary$`97.5%`
  )
})
```

### Results

```{r logistic_results}
logistic_res_df <- do.call(
  rbind,
  c(logistic_glm_reslist, logistic_ppd_reslist, logistic_psb_reslist)
)

logistic_res_df$est_ci <- sprintf(
  "%.3f (%.3f, %.3f)",
  logistic_res_df$estimate, logistic_res_df$lower, logistic_res_df$upper
)

wide <- reshape(
  logistic_res_df[, c("fitter", "borrowing", "variable", "est_ci")],
  direction = "wide",
  timevar = "fitter",
  idvar = c("borrowing", "variable"),
)
new_order <- order(wide$variable, wide$borrowing)
knitr::kable(wide[new_order, ], digits = 3, row.names = FALSE)
```

```{r logistic_plot}
logistic_res_df$borrowing_x <- logistic_res_df$borrowing +
  (as.numeric(factor(logistic_res_df$fitter)) - 3) / 100

ggplot(logistic_res_df, aes(x = borrowing_x, y = estimate, group = fitter, colour = fitter)) +
  geom_errorbar(aes(ymin = lower, ymax = upper)) +
  geom_point() +
  facet_wrap(~variable, scales = "free")
```

# Exponential models

Now we fit models with an exponentially distributed outcome. There is no censoring in this data set.
For `glm` we use `family = Gamma(link = "log")` and specify fixed `dispersion = 1` to fit a exponential model.
As before, the external control arm having weights (or power parameters)
equal to 0, 0.25, 0.5, 0.75, 1. The internal treated and control patients have weight = 1.
The model has a treatment indicator and two covariates, `eventtime ~ trt + cov1 + cov2`.

```{r surv_data}
sim_data_exp <- cbind(
  simsurv::simsurv(
    dist = "exponential",
    x = as.data.frame(example_matrix[, c("trt", "cov1", "cov2", "ext")]),
    betas = c("trt" = 1.3, "cov1" = 1, "cov2" = 0.1, "ext" = -0.4),
    lambdas = 5
  ),
  example_matrix[, c("trt", "cov1", "cov2", "ext")],
  censor = 0
)
```
```{r}
head(sim_data_exp)
```


```{r glm_exp}
## glm
glm_reslist <- lapply(c(0, 0.25, 0.5, 0.75, 1), function(w) {
  exp_glm <- glm(
    eventtime ~ trt + cov1 + cov2,
    data = sim_data_exp,
    family = Gamma(link = "log"),
    weights = ifelse(sim_data_exp$ext == 1, w, 1)
  )
  glm_summary <- summary(exp_glm, dispersion = 1)
  est <- -glm_summary$coefficients[, "Estimate"]
  lower <- est - 1.96 * glm_summary$coefficients[, "Std. Error"]
  upper <- est + 1.96 * glm_summary$coefficients[, "Std. Error"]

  data.frame(
    fitter = "glm",
    borrowing = w,
    variable = c("(Intercept)", "trt", "cov1", "cov2"),
    estimate = est,
    lower = lower,
    upper = upper
  )
})
```

```{r bayesppd_Exp, echo = FALSE, message = FALSE}
## BayesPPD
ppd_reslist <- lapply(c(0, 0.25, 0.5, 0.75, 1), function(w) {
  exp_ppd <- glm.fixed.a0(
    data.type = "Exponential",
    data.link = "Log",
    y = sim_data_exp[sim_data_exp$ext == 0, ]$eventtime,
    x = as.matrix(sim_data_exp[sim_data_exp$ext == 0, c("trt", "cov1", "cov2")]),
    historical = list(list(
      y0 = sim_data_exp[sim_data_exp$ext == 1, ]$eventtime,
      x0 = as.matrix(sim_data_exp[sim_data_exp$ext == 1, c("cov1", "cov2")]),
      a0 = w
    )),
    lower.limits = rep(-100, 5),
    upper.limits = rep(100, 5),
    slice.widths = rep(1, 5),
    nMC = 10000,
    nBI = 1000
  )[[1]]
  ci <- apply(exp_ppd, 2, quantile, probs = c(0.025, 0.975))
  data.frame(
    fitter = "BayesPPD",
    borrowing = w,
    variable = c("(Intercept)", "trt", "cov1", "cov2"),
    estimate = colMeans(exp_ppd),
    lower = ci[1, ],
    upper = ci[2, ]
  )
})
```

```{r bayesppd_Exp-no-eval, eval = FALSE}
## BayesPPD
ppd_reslist <- lapply(c(0, 0.25, 0.5, 0.75, 1), function(w) {
  exp_ppd <- glm.fixed.a0(
    data.type = "Exponential",
    data.link = "Log",
    y = sim_data_exp[sim_data_exp$ext == 0, ]$eventtime,
    x = as.matrix(sim_data_exp[sim_data_exp$ext == 0, c("trt", "cov1", "cov2")]),
    historical = list(list(
      y0 = sim_data_exp[sim_data_exp$ext == 1, ]$eventtime,
      x0 = as.matrix(sim_data_exp[sim_data_exp$ext == 1, c("cov1", "cov2")]),
      a0 = w
    )),
    lower.limits = rep(-100, 5),
    upper.limits = rep(100, 5),
    slice.widths = rep(1, 5),
    nMC = 10000,
    nBI = 1000
  )[[1]]
  ci <- apply(exp_ppd, 2, quantile, probs = c(0.025, 0.975))
  data.frame(
    fitter = "BayesPPD",
    borrowing = w,
    variable = c("(Intercept)", "trt", "cov1", "cov2"),
    estimate = colMeans(exp_ppd),
    lower = ci[1, ],
    upper = ci[2, ]
  )
})
```


```{r psborrow2_Exp, echo = FALSE, message = FALSE}
## psborrow2
psb_reslist <- lapply(c(0, 0.25, 0.5, 0.75, 1), function(w) {
  exp_psb2 <- create_analysis_obj(
    data_matrix = as.matrix(cbind(sim_data_exp, w = ifelse(example_matrix[, "ext"] == 1, w, 1))),
    covariates = add_covariates(c("cov1", "cov2"),
      priors = normal_prior(0, 100)
    ),
    borrowing = borrowing_details("Full borrowing",
      ext_flag_col = "ext",
    ),
    treatment = treatment_details("trt", normal_prior(0, 100)),
    outcome = outcome_surv_exponential("eventtime", "censor",
      baseline_prior = normal_prior(0, 1000),
      weight_var = "w"
    )
  )

  mcmc_exp_psb2 <- mcmc_sample(exp_psb2, chains = 1, verbose = FALSE, seed = 123)
  mcmc_summary <- mcmc_exp_psb2$summary(
    variables = c("alpha", "beta_trt", "beta[1]", "beta[2]"),
    mean,
    ~ quantile(.x, probs = c(0.025, 0.975))
  )

  data.frame(
    fitter = "psborrow2",
    borrowing = w,
    variable = c("(Intercept)", "trt", "cov1", "cov2"),
    estimate = mcmc_summary$mean,
    lower = mcmc_summary$`2.5%`,
    upper = mcmc_summary$`97.5%`
  )
})
```


```{r psborrow2_Exp-no-eval, eval = FALSE}
## psborrow2
psb_reslist <- lapply(c(0, 0.25, 0.5, 0.75, 1), function(w) {
  exp_psb2 <- create_analysis_obj(
    data_matrix = as.matrix(cbind(sim_data_exp, w = ifelse(example_matrix[, "ext"] == 1, w, 1))),
    covariates = add_covariates(c("cov1", "cov2"),
      priors = normal_prior(0, 100)
    ),
    borrowing = borrowing_details("Full borrowing",
      ext_flag_col = "ext",
    ),
    treatment = treatment_details("trt", normal_prior(0, 100)),
    outcome = outcome_surv_exponential("eventtime", "censor",
      baseline_prior = normal_prior(0, 1000),
      weight_var = "w"
    )
  )

  mcmc_exp_psb2 <- mcmc_sample(exp_psb2, chains = 1, verbose = FALSE, seed = 123)
  mcmc_summary <- mcmc_exp_psb2$summary(
    variables = c("alpha", "beta_trt", "beta[1]", "beta[2]"),
    mean,
    ~ quantile(.x, probs = c(0.025, 0.975))
  )

  data.frame(
    fitter = "psborrow2",
    borrowing = w,
    variable = c("(Intercept)", "trt", "cov1", "cov2"),
    estimate = mcmc_summary$mean,
    lower = mcmc_summary$`2.5%`,
    upper = mcmc_summary$`97.5%`
  )
})
```


### Results

Note: Wald confidence intervals are displayed here for `glm` for the exponential models.

```{r exp_results}
res_df <- do.call(rbind, c(glm_reslist, ppd_reslist, psb_reslist))

res_df$est_ci <- sprintf(
  "%.3f (%.3f, %.3f)",
  res_df$estimate, res_df$lower, res_df$upper
)

wide <- reshape(
  res_df[, c("fitter", "borrowing", "variable", "est_ci")],
  direction = "wide",
  timevar = "fitter",
  idvar = c("borrowing", "variable"),
)
new_order <- order(wide$variable, wide$borrowing)
knitr::kable(wide[new_order, ], digits = 3, row.names = FALSE)
```

```{r exp_plot}
res_df$borrowing_x <- res_df$borrowing +
  (as.numeric(factor(res_df$fitter)) - 3) / 100

ggplot(res_df, aes(x = borrowing_x, y = estimate, group = fitter, colour = fitter)) +
  geom_errorbar(aes(ymin = lower, ymax = upper)) +
  geom_point() +
  facet_wrap(~variable, scales = "free")
```
